---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CxGcensus

<!-- badges: start -->
<!-- badges: end -->

CxGcensus is an alternative *R* client to the [CELLxGENE census][]. It
emaphsizes use cases related to data discovery, and uses in-memory and
on-disk caches to reduce latency associated with repeated queries.

[CELLxGENE census]: https://chanzuckerberg.github.io/cellxgene-census/

## Installation

Install CxGcensus from [GitHub](https://github.com/) with:

``` r
## install.packages("devtools")
devtools::install_github("mtmorgan/CxGcensus")
```

## Example

Load the package

```{r example, messages = FALSE}
library(CxGcensus)
```

Discover datasets available in the census

```{r datasets}
datasets()
```

Summarize information about cells in the census

```{r summary_cell_counts}
summary_cell_counts() |>
    filter(category == "sex") |>
    select(
        organism, label,
        unique_cell_count, total_cell_count
    )
```

Learn about features (genes) in, e.g., `homo_sapiens` datasets in the census

```{r cars}
feature_data("homo_sapiens")
```

## Session information

This README was compiled with CxGcensus version
`r packageVersion("CxGcensus")`. Full session info is:

```{r}
sessionInfo()
```
