---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CxGcensus

<!-- badges: start -->
<!-- badges: end -->

CxGcensus is an alternative *R* client to the [CELLxGENE census][]. It
emaphsizes use cases related to data discovery, and uses in-memory and
on-disk caches to reduce latency associated with repeated queries.

[CELLxGENE census]: https://chanzuckerberg.github.io/cellxgene-census/

## Installation

Install CxGcensus from [GitHub](https://github.com/) with:

``` r
## install.packages("devtools")
devtools::install_github("mtmorgan/CxGcensus")
```

## Example

Load the package

```{r example, message = FALSE}
library(CxGcensus)
```

Discover datasets used to construct the census

```{r datasets}
datasets()
```

`dataset_id` and `collection_id` can be used in the [cellxgenedp][]
package for programmatic retrieval and visualization of individual
datasets.

The organisms present in the census (use these in calls to
`feature_data()` or `observation_data()`) are available with

```{r}
census_names()
```

Summarize information about cells in the census

```{r summary_cell_counts}
summary_cell_counts() |>
    filter(category == "sex") |>
    select(
        organism, label,
        unique_cell_count, total_cell_count
    )
```

Learn about features (genes) in, e.g., `homo_sapiens` datasets in the census

```{r cars}
feature_data("mus_musculus")
```

Observation (cell) data annotations are available with
`observation_data()`. The strategy adopted is to download all the data
to a local [duckdb][] database. Download can be time-consuming, but
the results are 'memoised' so subsequent calls are more-or-less
instantaneous.

```{r observation_data}
mus <- observation_data("mus_musculus")
mus
```

The columns available for cross-dataset analysis are

```{r}
mus |>
    colnames()
```

It is straight-forward to summarize or query the data, e.g., for the
number of different types of assays represented, or for studies of
diabetes.

```{r}
mus |>
    count(assay, sort = TRUE)
mus |>
    filter(grepl("diabetes", disease)) |>
    count(disease, sex, tissue)
```

The `soma_joinid` in the tibbles returned by `feature_data()` and
`observation_data()` are important in retrieving counts.

[cellxgenedp]: https://bioconductor.org/packages/cellxgenedp
[duckdb]: https://cran.r-project.org/package=duckdb

## Session information

This README was compiled with CxGcensus version
`r packageVersion("CxGcensus")`. Full session info is:

```{r}
sessionInfo()
```
