---
title: "C. Comparison with CuratedAtlasQuery and cellxgene.census"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{C. Comparison with CuratedAtlasQuery and cellxgene.census}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CxGcensus)
```

# CuratedAtlasQuery

## An example query

The *R* package [CuratedAtlasQueryR][] contains independently curated
CELLxGENE data sets. Start by installing the package

```{r, eval = FALSE}
remotes::install_github("stemangiola/CuratedAtlasQueryR")
```

Load the library

```{r CuratedAtlasQueryR, message = FALSE}
library(CuratedAtlasQueryR)
```

Retrieve the (large) metadata file.

```{r, message = FALSE}
metadata <- get_metadata()
```

Query for cells from samples with 'African' ancestry, lung parenchyma
tissue, assayed with 10x technology, and with cell type containing the
string 'CD4'.

```{r}
single_cell_counts <-
    metadata |>
    dplyr::filter(
        ethnicity == "African" &
        stringr::str_like(assay, "%10x%") &
        tissue == "lung parenchyma" &
        stringr::str_like(cell_type, "%CD4%")
    )
single_cell_counts
single_cell_counts |>
    dplyr::count(dataset_id)
```

Retrieve the cells as a single-cell experiment

```{r, message = FALSE}
system.time({
    sce_caqr <-
        single_cell_counts |>
        get_single_cell_experiment()
})
sce_caqr
```

The `assay()` data is a `DelayedArray` matrix of counts; the nature of
the count is described in the `sce_caqr$x_normalization` column, but
in this case is not informative.

```{r}
class(SummarizedExperiment::assay(sce_caqr))
range(as(SummarizedExperiment::assay(sce_caqr), "sparseMatrix"))
sce_caqr |>
    SummarizedExperiment::colData() |>
    as.data.frame() |>
    dplyr::count(x_normalization)
```

[CuratedAtlasQueryR]: https://stemangiola.github.io/CuratedAtlasQueryR

## The query with CxGcensus

The first time `observation_data()` is run, it creates a local
database for efficient access. The database is created in 'chunks' to
moderate memory consumption. The initial code here makes sure that
memory consumption is modest.

```{r}
description <- cellxgene.census::get_census_version_description("stable")
uri <- description$soma.uri
tiledbsoma_ctx <-
    cellxgene.census::new_SOMATileDBContext_for_census(description)
## chunk size, from Dirk Eddelbuettel, via slack
tiledbsoma_ctx$set("soma.init_buffer_bytes", as.character(1024**3 / 10))

homo <- observation_data(
    "homo_sapiens", uri = uri, tiledbsoma_ctx = tiledbsoma_ctx
)
```

```{r}
obs_query <-
    observations |>
    filter(
        self_reported_ethnicity == "African", # 431882 cells
        tissue == "lung parenchyma",          # 189662 cells
        assay %LIKE% "%10x%",                 # 189662 cells
        cell_type %LIKE% "%CD4%"              # 3371
    )

obs_query |>
    count(dataset_id, is_primary_data)

obs_query_exact <-
    obs_query |>
    filter(dataset_id == "066943a2-fdac-4b29-b348-40cede398e4e") |>
    collect()
```

```{r, eval = FALSE}
feat_query <-
    feature_data("homo_sapiens")
```

```{r, eval = FALSE}
feat_query_approx <-
    feature_data("homo_sapiens") |>
    ## not all features are present -- SYMBOL is a very poor choice
    ## for identifier
    filter(feature_name %in% rownames(sce_caqr))
```

```{r, eval = FALSE}
system.time({
    sce_cxg <-
        single_cell_experiment("homo_sapiens", feat_query, obs_query_exact)
})
```

# cellxgene.census

[cellxgene.census][]


The following is based on the cellxgene census [R Quick Start][]. Load
the package.

```{r, message = FALSE, eval = FALSE}
library(cellxgene.census)
```

Open a connection to the census as a 'soma' object, and retrieve a
reference to the 'obs'ervation data.

```{r, message = FALSE, eval = FALSE}
soma <- open_soma(census_version = "2023-05-15")
observation_data <- soma$get("census_data")$get("homo_sapiens")$get("obs")
```

Discover available values for filtering

```{r, eval = FALSE}
observation_data$colnames()
```

Formulate a filter using the syntax of *cellxgene.census*

```{r, eval = FALSE}
value_filter <- paste(
    "self_reported_ethnicity == 'African'",
    "tissue == 'lung parenchyma'",
    "cell_type == 'CD4-positive, alpha-beta T cell'",
    ## assay? any of "10x 5' v1", "10x 3' v2", "10x 3' v3"
    sep = " & "
)
```

Read the metadata associated with the query

```{r, eval = FALSE}
system.time({
    observations <-
        observation_data$read(value_filter = value_filter)$concat() |>
        as.data.frame()                             # actually, tibble
})
observations
observations |>
    dplyr::count(dataset_id)
```

There are 1800 more cells and a second dataset; this is not because
the filter did not include an 'assay' component; perhaps `soma`
includes datasets that were not available when *CuratedAtlasQueryR*
independently harmonized data, `suspension_type`, or to a
hard-to-understand (for me!)  classification as primary data
(*CuratedAtlasQueryR* return only dataset
`066943a2-fdac-4b29-b348-40cede398e4e`)

```{r, eval = FALSE}
observations |>
    count(dataset_id, suspension_type, is_primary_data)
```

[cellxgene.census][] does not support representaton as
SingleCellExperiment; retrieve the Seurat object corresponding to the
observations we are interested in.

```{r, eval = FALSE}
system.time({
    seurat <- cellxgene.census::get_seurat(
        census = soma,
        organism = "Homo sapiens",
        obs_value_filter = value_filter
    )
})
```

[cellxgene.census]: https://chanzuckerberg.github.io/cellxgene-census/
[R Quick Start]:  https://chanzuckerberg.github.io/cellxgene-census/cellxgene_census_docsite_quick_start.html#r-quick-start
